\section{Relational Algebra}

% ----------------------------------------------------------
\begin{frame}{What is Relational Algebra?}
  Relational algebra is a \textbf{formal, procedural query language} for
  relational databases.  Unlike declarative languages (where you state
  \emph{what} you want), relational algebra requires you to specify
  \emph{how} to retrieve data by composing a sequence of well-defined
  operations.  It forms the mathematical backbone of SQL and every other
  relational query language.

  \begin{block}{Core Properties}
    \begin{itemize}
      \item Every operation takes one or two \emph{relations} as input and
            produces exactly one \emph{relation} as output — the
            \textbf{closure property}.  This means operations can be
            composed freely.
      \item Provides the theoretical foundation for SQL; every SQL
            \texttt{SELECT} statement can be translated into a relational
            algebra expression.
      \item By combining operations arbitrarily, queries of any complexity
            can be expressed.
    \end{itemize}
  \end{block}

  \begin{block}{Named Operators}
    Union $\cup$, Intersection $\cap$, Difference $\setminus$, Cartesian
    Product $\times$, Division $\div$, Restriction $\sigma$, Projection
    $\pi$, Rename $\rho$, Join $\bowtie$
  \end{block}
\end{frame}

% ----------------------------------------------------------
\begin{frame}{Overview of All Operators}
  The table below lists all operators covered in this lecture.  Unary
  operators act on a single relation; binary operators combine two
  relations.  Join operators are a special family of binary operators and
  are treated separately in later frames.

  \vspace{0.4em}
  \begin{center}
    \footnotesize
    \begin{tabular}{llll}
      \hline
      \textbf{Type} & \textbf{Symbol} & \textbf{Operator} & \textbf{Description} \\
      \hline
      Binary & $\cup$           & Union              & All tuples from both relations, no duplicates \\
      Binary & $\cap$           & Intersection       & Only tuples common to both \\
      Binary & $\setminus$      & Difference         & Tuples in $R_1$ but not in $R_2$ \\
      Binary & $\times$         & Cartesian Product  & Every combination of tuples \\
      Binary & $\div$           & Division           & Tuples in $R_1$ associated with ALL tuples in $R_2$ \\
      Unary  & $\sigma$         & Restriction        & Filter rows by condition \\
      Unary  & $\pi$            & Projection         & Select specific columns \\
      Unary  & $\rho$           & Rename             & Rename relation or attributes \\
      Unary  & $\delta$         & Deduplication      & Remove duplicate tuples \\
      Unary  & $\gamma$         & Grouping/Aggregation & Group and aggregate \\
      Unary  & $\tau$           & Sorting            & Sort tuples \\
      Join   & $\bowtie$        & Join               & Combine related tuples from two relations \\
      \hline
    \end{tabular}
  \end{center}
\end{frame}
