% ---------------------------------------------------------------------------
\begin{frame}{4NF --- The Problem: Independent Multi-Valued Facts \textbf{(!)}}
  \begin{block}{Motivation}
    After BCNF, a relation can still contain \textbf{redundancy} caused by
    independent multi-valued facts being stored in the same table.
  \end{block}

  Consider a customer who has \textbf{multiple phone numbers} AND \textbf{multiple
  addresses} --- both \emph{independent} of each other.

  \vspace{6pt}
  \begin{center}
  \scriptsize
  \begin{tabular}{lll}
    \toprule
    \textbf{CustomerCode} & \textbf{Phone} & \textbf{Address} \\
    \midrule
    C001 & 0711-123 & Hauptstr.\ 1, Stuttgart \\
    C001 & 0711-123 & Bahnhofstr.\ 5, Stuttgart \\
    C001 & 0152-456 & Hauptstr.\ 1, Stuttgart \\
    C001 & 0152-456 & Bahnhofstr.\ 5, Stuttgart \\
    \bottomrule
  \end{tabular}
  \end{center}
  \normalsize

  \vspace{4pt}
  \begin{alertblock}{The Problem: Cartesian Product}
    Every phone number is paired with every address --- even though they are
    \textbf{independent} facts.  2 phones $\times$ 2 addresses = 4 rows.
    Adding one new phone number requires inserting \emph{two} rows.
  \end{alertblock}
\end{frame}

% ---------------------------------------------------------------------------
\begin{frame}{4NF --- Definition and Fix \textbf{(!)}}
  \begin{block}{Definition \textbf{(!)}}
    A relation is in \textbf{4NF} if it is in BCNF and contains no
    \textbf{non-trivial multi-valued dependencies (MVDs)}.

    An MVD $A \twoheadrightarrow B$ is \emph{non-trivial} if $B \not\subseteq A$
    and $A \cup B \neq R$.
  \end{block}

  \begin{block}{The MVDs in Our Example}
    $\texttt{CustomerCode} \twoheadrightarrow \texttt{Phone}$ \quad
    (customer has a set of phones)\\
    $\texttt{CustomerCode} \twoheadrightarrow \texttt{Address}$ \quad
    (customer has a set of addresses)\\
    These two facts are \textbf{independent} of each other --- 4NF violation.
  \end{block}

  \begin{block}{Fix: One Table Per Independent Multi-Valued Fact}
    \begin{itemize}
      \item \texttt{CustomerPhone(\underline{CustomerCode, Phone})}
      \item \texttt{CustomerAddress(\underline{CustomerCode, Address})}
    \end{itemize}
    Now each fact lives in its own table.  No spurious Cartesian product.
  \end{block}
\end{frame}

% ---------------------------------------------------------------------------
\begin{frame}{5NF --- Project-Join Normal Form \textbf{(!)}}
  \begin{block}{Definition \textbf{(!)}}
    A relation is in \textbf{5NF} (also called \emph{PJNF}) if it is in 4NF
    and cannot be further \textbf{losslessly} decomposed.
  \end{block}

  \begin{block}{Key Concepts}
    \begin{itemize}
      \item \textbf{Lossless:} $R_1 \rajoin R_2 \rajoin \cdots = R$ exactly
            (no spurious tuples added).
      \item A 2-table split that produces extra rows when rejoined is
            \emph{lossy} --- a 3-table decomposition may be needed.
      \item 5NF deals with \textbf{join dependencies} (generalisations of MVDs).
    \end{itemize}
  \end{block}

  \begin{alertblock}{Practical Note}
    Rarely implemented in practice.  Most practitioners stop at 3NF or BCNF.
    Verifying 5NF requires checking all possible join dependencies ---
    computationally expensive and usually unnecessary.
  \end{alertblock}
\end{frame}

% ---------------------------------------------------------------------------
\begin{frame}{Even Higher Normal Forms \textbf{(!)}}
  Beyond 5NF, further normal forms have been defined mostly in academic research.

  \begin{itemize}
    \item \textbf{6NF:} Every relation contains the PK and \emph{at most one}
          non-key attribute.  Useful for \textbf{temporal databases} where
          attributes change independently over time.
          (Date, Darwen, Lorentzos, 2002)
    \item \textbf{EKNF (Elementary Key NF):} A stricter variant of 3NF;
          rules out certain anomalous 3NF schemas.
    \item \textbf{ETNF (Essential Tuple NF):} Stricter than 4NF; eliminates
          redundancies that 4NF does not address.
    \item \textbf{DKNF (Domain-Key NF):} Every integrity constraint is a logical
          consequence of domain and key constraints.  Considered the
          ``ultimate'' normal form --- all anomalies guaranteed absent.
  \end{itemize}

  \begin{alertblock}{Practical Guidance}
    Most of these forms are \textbf{academic}.  In practice:
    \begin{itemize}
      \item Aim for \textbf{3NF minimum} for all production schemas.
      \item \textbf{BCNF} is ideal when dependency preservation can be
            verified externally or enforced by application logic.
    \end{itemize}
  \end{alertblock}
\end{frame}

% ---------------------------------------------------------------------------
\begin{frame}{Normal Forms: The Big Picture \textbf{(!)}}
  \begin{center}
  \scriptsize
  \begin{tabular}{lcccccccc}
    \toprule
    \textbf{Constraint} & \textbf{UNF} & \textbf{1NF} & \textbf{2NF} & \textbf{3NF} & \textbf{BCNF} & \textbf{4NF} & \textbf{5NF} & \textbf{6NF} \\
    \midrule
    Atomic values (no lists)           & & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
    No partial FDs                     & & & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
    No transitive FDs                  & & & & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
    Every FD LHS is a superkey         & & & & & \checkmark & \checkmark & \checkmark & \checkmark \\
    No non-trivial MVDs                & & & & & & \checkmark & \checkmark & \checkmark \\
    No lossy 2-table decomposition     & & & & & & & \checkmark & \checkmark \\
    PK + at most one non-key attr.     & & & & & & & & \checkmark \\
    \bottomrule
  \end{tabular}
  \end{center}
  \normalsize

  \vspace{6pt}
  \begin{block}{How to Read This Table}
    A \checkmark{} means the constraint is \emph{guaranteed} by that normal form
    and all higher ones.  Each NF strictly extends its predecessor.
  \end{block}

  \begin{block}{Practical Target}
    \textbf{3NF} for most systems --- \textbf{BCNF} when additional precision is
    needed and FD loss is acceptable.
  \end{block}
\end{frame}
